---
title: "project2-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project2-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MATH5793BANH)
```


```{r}
stock <- read.table("C:\\Users\\kevin\\OneDrive\\Documents\\MATH5793\\Project2\\T8-4.DAT")
colnames(stock) <- c("JP-Morgan", "Citibank","Wells-Fargo","Royal-Dutch-Shell","Exxon-Mobil")
head(stock)
```



# Function for part A

```{r}
pcaCalc <- function(data){
  cov <- cov(data)
  eig <- eigen(cov)
  
  n <- dim(cov)[1]
  p <- dim(cov)[2]
  
  rho <- matrix(, nr = n, nc = p)
  
  for(i in 1:n){
    for(j in 1:p){
      rho[i,j] = eig$vectors[i,j]*sqrt(eig$values[i])/sqrt(cov[j,j])
    }
  }
  
  return(list(S = cov, sample.princ.comp = eig$vectors, total.variance = eig$values, correlation = rho))
}
```




# a)
```{r}
# Sample covariance matrix
pcaCalc(stock)$S


# Correlation matrix of r(yi,xk)
pcaCalc(stock)$sample.princ.comp
```



# b)
```{r}
sum = 0

for(i in 1:3){
  sum = sum + pcaCalc(stock)$total.variance[i]
}

sum/sum(pcaCalc(stock)$total.variance)
```


90% of the total variability in stocks rate of return is explained by the first 3 principal components -- JP Morgan, Citibanks, and Wells-Fargo


# Function for part C
```{r}
bonf.cis <- function(x, n, p,alpha){
  # x is a vector of variances
  # n is the number of rows
  # p is the number of columns
  # alpha is the significance level
  
  for(i in 1:length(x)){
    zcalc <- qnorm(1-alpha/(2*p))
    lower.lim <- x[i]/(1 + zcalc*sqrt(2/n))
    upper.lim <- x[i]/(1 - zcalc*sqrt(2/n))
    print(paste(lower.lim, "<= lambda", i, " <= ", upper.lim))
  }
}
```




# c) 
```{r}
n <- dim(stock)[1]
p <- dim(stock)[2]
var3 <- c(pcaCalc(stock)$total.variance[1],pcaCalc(stock)$total.variance[2],pcaCalc(stock)$total.variance[3])
bonf.cis(var3,n,p,.10)
```



# d)

```{r}
v <- pcaCalc(stock)$total.variance
(v[1] + v[2])/(sum(v))
```

The stocks rates-of-return data can be summarized by the first 2 components (JP Morgan and Citibank) because given our calculations in b and c. In particular, in b we showed that the first 3 principal components explained 90% of the variability in our data. However, in c we know that the confidence intervals get wider as our $\lambda s$ get wider. It's apparent that the Bonferroni confidence intervals for the first two $\lambda s$ are much wider than the third, in addition to our calculation in d), the first two components explain 80% of the variability in our data thus we can reduce the data to just 2 dimensions.
