---
title: "project2-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project2-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MATH5793BANH)
```


```{r}
stock <- read.table("C:\\Users\\kevin\\OneDrive\\Documents\\MATH5793\\Project2\\T8-4.DAT")
colnames(stock) <- c("JP-Morgan", "Citibank","Wells-Fargo","Royal-Dutch-Shell","Exxon-Mobil")
head(stock)
```



# Function for part A

```{r}
pcaCalc <- function(data){
  cov <- cov(data)
  eig <- eigen(cov)

  n <- dim(cov)[1]
  p <- dim(cov)[2]

  rho <- matrix(, nrow = n, ncol = p)

  for(i in 1:n){
    for(j in 1:p){
      rho[i,j] = eig$vectors[i,j]*sqrt(eig$values[i])/sqrt(cov[j,j])
    }
  }

  total.sample.var <- sum(eig$values)

  return(list(S = cov, sample.princ.comp = eig$vectors, variance = eig$values, total.variance = total.sample.var, correlation = rho))
}
```




# a)
```{r}
# Sample covariance matrix
pcaCalc(stock)$S


# Correlation matrix of r(yi,xk)
pcaCalc(stock)$sample.princ.comp
```

Principal components are $\hat{y_i} = e_i'x$ where i is the column of the eigen vector output and e is an eigenvector. Thus for $\hat{y_1}$ and $\hat{y_2}$

$$\hat{y_1} = .223x_1 + .307x_2 + .155x_3 + .639x_4 + .651x_5$$

$$\hat{y_2} = .625x_1 + .570x_2 + .345x_3 -.248x_4 - .322x_5$$

$$\hat{y_3} = .326x_1 - .250x_2 - .038x_3 - .642x_4 + .646x_5$$

$$\hat{y_4} = .663x_1 - .414x_2 - .497x_3 + .309x_4 -.216x_5$$


$$\hat{y_5} = .118x_2 - .590x_2 + .780x_3 + .148x_4 - .094x_5$$


# b)
```{r}
sum = 0

for(i in 1:3){
  sum = sum + pcaCalc(stock)$variance[i]
}

sum/pcaCalc(stock)$total.variance
```


90% of the total variability in stocks rate of return is explained by the first 3 principal components -- JP Morgan, Citibanks, and Wells-Fargo


# Function for part C
```{r}
bonf.andgir.cis <- function(data, q, alpha){
  n = dim(data)[1]
  p = dim(data)[2]
  
  eigen.val <- c()
  for(i in 1:q){
    eigen.val[i] = eigen(cov(data))$value[i]
  }
  
  bonf.ci <- list()
  for(i in 1:q){
    zcalc <- qnorm(1-alpha/(2*q))
    lower.lim <- eigen.val[i]/(1 + zcalc*sqrt(2/n))
    upper.lim <- eigen.val[i]/(1 - zcalc*sqrt(2/n))
    bonf.ci[[i]] = c(lower.lim, upper.lim)
  }
  
  return(bonf.ci)
}
```




# c) 
```{r}
bonf.andgir.cis(stock,3,alpha = .10)
```



# d)

```{r}
v <- pcaCalc(stock)$variance
(v[1] + v[2])/(sum(v))
```

The stocks rates-of-return data can be summarized by the first 2 components (JP Morgan and Citibank) because given our calculations in b and c. In particular, in b we showed that the first 3 principal components explained 90% of the variability in our data. However, in c we know that the confidence intervals get wider as our $\lambda s$ get wider. It's apparent that the Bonferroni confidence intervals for the first two $\lambda s$ are much wider than the third, in addition to the first two components explain 80% of the variability in our data thus we can reduce the data to just 2 dimensions.



